Novel lattice Boltzmann method for simulation of strongly shear thinning viscoelastic fluids

### 2024-02-20 NovelLBM.tex
plots

To illustrate our novel LBM algorithm, we consider two prototypical cell carrier fluids.
The first is an alginate solution which is often used in biofabrication but can also serve as a common food additive.
Our solution contains \SI{4}{\percent} alginate dissolved in PBS and was measured in a commercial plate-plate rheometer (see appendix~\ref{app:experiment} for experimental details).
The shear-dependent viscosity $\eta$ and the first normal stress difference $N_1$ are shown as function of shear rate in figure~\ref{fig:fit_alginate}.
We note the very high zero-shear viscosity of more than \SI{10000}{\milli\pascal\second}.
Bioinks in biofabrication are in general characterized by such high viscosities to guarantee stability of the printed construct.
At the same time, during the actual printing process when the material is flowing, the viscosity should be as low as possible to avoid excessive mechanical stress on the embedded cells.
This motivates the use of highly shear-thinning viscoelastic fluids in biofabrication.
	\begin{figure}[H]
		\centering
		\adjincludegraphics[trim=0 0 0 0, clip, scale=0.5]{assets/plots/fit_alginate_eta.eps}
		\adjincludegraphics[trim=0 0 0 0, clip, scale=0.5]{assets/plots/fit_alginate_N.eps}
		\caption{Rheology data for alginate solution from 21 measurements of the viscosity (a) and normal stress (b).
				Orange lines show a fit of the PTT rheological model. \SG{ larger font in figures }}
		\label{fig:fit_alginate}
	\end{figure}\todo{larger fonts in all figures}

The second cell carrier fluid modeled in this work is a methyl cellulose solution which is commonly used in real-time-deformation-cytometry (RT-DC) experiments.
RT-DC experiments allow very high-throughput measurements of cell mechanical properties and have been used successfully for disease detection by screening blood samples.
We extract our data from \cite{buyukurganci2023} where the shear rheology of methyl cellulose was measured using a plate-plate and cone-plate rheometer.
Three concentrations of \SI{0.49}{\percent}, \SI{0.59}{\percent} and \SI{0.83}{\percent} methyl cellulose in PBS were investigated, and their rheology is shown in figure~\ref{fig:fit_mc}.
Even at the highest concentration the zero-shear viscosity is two orders of magnitude lower than for the considered alginate solution.
	\begin{figure}[H]
		\centering
		\includegraphics[width=7.8cm]{assets/plots/fit_mc_eta.eps}
		\includegraphics[width=7.8cm]{assets/plots/fit_mc_N.eps}
		\caption{Rheology data of a methyl cellulose solution from \cite{buyukurganci2023}.
				Orange lines show a fit of the PTT rheological model. \SG{ include legend into figure. Maybe give color coding only in caption? }\RK{meaning ony one legend for both plots?}\SG{ larger font in figures }}
		\label{fig:fit_mc}
	\end{figure}

@@

The PTT model is specified by a differential equation for the polymer stress $\ma{\tau}$:
	\begin{equation}
		\overset{\nabla}{\ma{\tau}} = -\xi\left(\ma{\tau}\cdot\ma{D} + \ma{D}\cdot\ma{\tau}\right) - e^{\frac{\epsilon\lambda}{\eta_\text{p}}\Tr\ma{\tau}}\frac{\ma{\tau}}{\lambda} + 2\frac{\eta_\text{p}}{\lambda}\ma{D}
		\label{eq:polymerStress}
	\end{equation}
where $\ma{D}$ is the strain rate tensor.
\SG{ check that PTT is given in this form in \cite{phan-thien1977,phan-thien1978} or cite other source }
The polymer viscosity is denoted $\eta_\text{p}$ and $\lambda$ is the corresponding relaxation time.
$\epsilon$ is the so-called extensibility parameter which controls at what shear rates thinning takes place and $\xi$ describes the slip in the polymer network which we will set to zero in the following.
The upper convected derivative of an arbitrary tensor $\ma{A}$ is defined as follows:
	\begin{equation}
		\overset{\nabla}{\ma{A}} = \frac{\Dif\ma{A}}{\Dif t} - \left(\left(\nabla\vec{u}\right)^\mathrm{T}\cdot\ma{A} + \ma{A}\cdot\left(\nabla\vec{u}\right)\right)
	\end{equation}
where $\vec{u}$ is the fluid velocity and $\left(\nabla\vec{u}\right)_{ij} = \nabla_i\vec{u}_j$.
The material derivative is
	\begin{equation}
		\frac{\Dif\ma{A}}{\Dif t} = \frac{\partial\ma{A}}{\partial t} + \vec{u}\cdot\nabla\ma{A}.
	\end{equation}
Finally, the total fluid stress $\ma{\sigma}$ is the sum of the polymer stress $\ma{\tau}$ and a solvent stress due to the Newtonian solvent with viscosity $\eta_\text{s}$:
	\begin{align}
		\ma{\sigma} &= \ma{\tau} + 2\eta_\text{s}\ma{D} 		\label{eq:totalStress}
	\end{align}

@@

We introduce the ratio between the zero-shear and the infinite-shear (i.e.\ solvent) viscosity as\todo{nope $\eta_p$ is not zero shear viscosity!}
\begin{align}
	R_\eta = \frac{\eta_\text{p}}{\eta_\text{s}}		\label{eq:R_eta}
\end{align}
which will turn out to be an important parameter to determine numerical stability of LB simulations.
\SG{ wo don't actually show that $R_ \eta  $ determines stability? }\RK{we do not. We could do a plot with $alpha_s$ decreasing until the sim breaks if you want}
This ratio can similarly be defined for other models such as FENE-P or the inelastic, but shear-thinning Carreau-Yasuda model \cite{carreau1972, yasuda1979}.
For our alginate solution we obtain $R_\eta = \num{48.2\pm0.4e3}$, indicating that the solution viscosity at very low shear rates is more than four orders of magnitude larger than the solvent viscosity which represents a severe challenge to standard numerical algorithms.
That this characteristic is not unique to alginate, but is in fact a generic feature of cell carrier fluids used in bioprinting is shown in table~\ref{tab:bioinks} where rheological data from various literature sources have been compiled into a single format to determine $R_\eta$.
The \SI{0.49}{\percent} methyl cellulose solution used in RT-DC measurements, in contrast, features a much lower value of $R_\eta = \num{18.7\pm0.4}$.

@@

The spatially resolved errors are shown in figure~\ref{fig:2DvelErr} in the appendix.
Averaging over the channel diameter, we obtain an L2 error of \num{9e-4} and \SG{ XX } for alginate and methyl cellulose, respectively.
These conclusions remain robust even when varying the pressure gradient, and thus the flow velocity, over many orders of magnitude as shown in figure~\ref{fig:pressureErr}~(a) of the appendix.\RK{This is the worst case error for MC and it is tiny "remain robust" is too weak here}

	\begin{figure}[H]
		\centering
		\adjincludegraphics[trim=0 0 0 0, clip, scale=0.5]{assets/plots/2DAlginateU.eps}
		\adjincludegraphics[trim=0 0 0 0, clip, scale=0.5]{assets/plots/2DParameterStudyUWorst.eps}
		\caption{(a) Velocity profile of alginate in planar channel flow. \SG{use $y$ instead of $r$ in 2D}.
		The time evolution of the center-line velocity is shown in the inset \SG{ to do }.
		(b) Velocity profile for methyl cellulose.
		\SG{ make these plots for different $ \alpha _ \mathrm{s}$, either include here or put into appendix }}\RK{Why? Is identical down to fp32 precision (see appendix C)}
		\label{fig:2Dvel}
	\end{figure}\todo{maybe try if it is this identical for different $\alpha_s$ first}

We proceed to evaluate the viscosity shuffling method for a cylindrical Poiseuille flow.
The semi-analytical solution for this situation can be derived similarly as for the planar case and is given in appendix~\ref{app:semianalyticalPoiseuille}.
Numerically, the cylindrical situation is slightly more complex due to discretization errors ("staircase effect") when mapping the rounded channel walls onto the rectilinear LBM grid \cite{Lehmann.20229da}.
We use the same material parameters as above, but double the pressure gradient to \SG{ XX }. \SG{ why? } \RK{where analytical solutions exist in Poiseuille geometry the 3D solution differs from the 2D solution by a factor 2. See the velocities in our sims being the same with this scaling}
The channel has a diameter of \SI{20}{\micro\meter}\SG{ XX }\RK{done} and is discretized with 41 lattice points.
Again, we initialize the polymers in the unstretched state for methyl cellulose and in the stretched state for alginate.
The resulting velocity profile can be seen in figure~\ref{fig:3Dvel} and shows almost equally good agreement as for the planar case with an L2 error of \num{0.014} and \SG{ XX }, respectively.
The spatially resolved error in figure~\ref{fig:3DvelErr} of the appendix shows that this larger average is indeed mainly attributable to the staircase effect as the error near the walls increases significantly.
The robustness against pressure gradient changes is shown in figure~\ref{fig:pressureErr}~(b) of the appendix.

	\begin{figure}[H]
		\centering
		\adjincludegraphics[trim=0 0 0 0, clip, scale=0.5]{assets/plots/3DAlginateU.eps}
		\adjincludegraphics[trim=0 0 0 0, clip, scale=0.5]{assets/plots/3DParameterStudyUWorst.eps}
		\caption{(a) Velocity profile of alginate in cylindrical channel flow.
		The time evolution of the center-line velocity when starting from a pre-stretched configuration (see main text) is shown in the inset.
		(b) Velocity profile for methyl cellulose.
				\SG{ make these plots for different $ \alpha _ \mathrm{s}$, either include or put into Si }}
		\label{fig:3Dvel}
	\end{figure}\todo{see above}

@@

	\begin{figure}[H]
		\centering
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/malaspinas2010.eps}
		\caption{L2 error as defined in equation \ref{eq:l2error} from Malaspinas et al.\cite{malaspinas2010} (dots) compared to our algorithm (x) as a function of grid resolution.
		Blue and green crosses are not visible as they overlap exactly with the orange and red crosses, respectively, showing that the accuracy of our method is independent of $\mathrm{Wi}$ in the studied parameter range.}
		\label{fig:malaspinas2010}
	\end{figure}

As can be seen in figure~\ref{fig:malaspinas2010}, the accuracy of our method does not depend on the Weissenberg Number and is generally somewhat better than the method of \cite{malaspinas2010}.
We note a certain dependence of the accuracy on $R_\nu$ which we consider nevertheless acceptable given the overall smallness of the error.
Finally, we find that the error of our method decreases with lattice resolution to the power of \SG{ give number, show dashed line in plot }.\RK{They see diminishing returns we don't}

@@

The first application stems from bioprinting where the bioink is pushed through a conical nozzle \cite{possl2021, Poologasundarampillai.2021, oconnell2022}.
The nozzle radius is \SI{1.8}{\milli\meter} at the inlet and \SI{200}{\micro\meter} at the outlet.
The nozzle has a length of \SI{18}{\milli\meter}.
Due to the difference in inlet/outlet radii, the system cannot be modeled with periodic boundary conditions, and we use inflow/outflow boundary conditions as described in section~\ref{sec:Methods} instead.
The required values for velocity and polymer conformation are obtained from simulations of pipes of the respective radii and a fixed volume flow of $\approx\SI{3}{\micro\liter\per\second}$.
The resulting flow field can be seen in figure \ref{fig:Nozzle_v}~(a).
To quantify the total stress as a single number, we use the von Mises stress which is defined as \cite{Mueller.2023}:
\begin{align}
	\sigma_\text{vM} = \sqrt{\frac{1}{2}\left[\left(\sigma_1 - \sigma_2\right)^2 + \left(\sigma_2 - \sigma_3\right)^2 + \left(\sigma_3 - \sigma_1\right)^2\right]}.
\end{align}
It can be seen that the highest stresses arise at the nozzle exit in direct vicinity to the wall.
This means that even for fast flows, cells flowing close to the center experience less stress than cells flowing near the walls in agreement with earlier results in Newtonian fluids \cite{Mueller.2023}.
	\begin{figure}[H]
		\centering
		\adjincludegraphics[trim=0 0 0 0, clip, scale=0.5]{assets/plots/Nozzle_v.eps}
		\adjincludegraphics[trim=0 0 0 0, clip, scale=0.5]{assets/plots/Nozzle_sigma_vM.eps}
		\caption{Velocity profile (a) and von Mises stress (b) for methyl cellulose in a conical nozzle as used for bioprinting. \SG{ draw nozzle contour with black lines }}\RK{looks terrible}
		\label{fig:Nozzle_v}
	\end{figure}

@@

 	\begin{figure}[H]
		\centering
		\adjincludegraphics[trim=0 0 0 0, clip, scale=0.5]{assets/plots/RT-DC_v.eps}
		\adjincludegraphics[trim=0 0 0 0, clip, scale=0.5]{assets/plots/RT-DC_tau.eps}
		\caption{Velocity profile (a) and shear stress (b) for methyl cellulose in a typical RT-DC geometry. \SG{ draw channel contour with black lines }
		\SG{ change to von Mises instead of $ \tau _{ 12 }$? }}
		\label{fig:RT-DC}
	\end{figure}

@@

	\begin{figure}[H]
		\centering
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/cellInShear_sigma_vM.eps}
		\caption{The von Mises stress around a cell in viscoelastic shear flow.}
		\label{fig:cellInShear_vM}
	\end{figure}

@@

\section{Semi-analytical solution for Poiseuille flows of PTT fluids with solvent viscosity}
\label{app:semianalyticalPoiseuille}
\SG{ make a self-contained derivation which is understandable even without reading Oliveira \\}
\SG{ use Gl.~(\ref{eq:polymerStress}) as starting point and end with Poiseuille solution, but also shear solution Gl.~(\ref{eq:shearN}, \ref{eq:shearEta}) \\}
Our derivation is analogous to Oliveira\cite{oliveira1999}. To account for the added viscous stress due to $\eta_\text{s}$, equilibrium is reached at lower polymer stresses. Therefore, Oliveiras equation (7) needs to be modified as follows.
	\begin{equation}
		\tau_{xy} = \partial_xp\frac{r}{2^j} - \eta_\text{s}\dot{\gamma}
	\end{equation}
	Where the second term is added due to the presence of the solvent viscosity. This term also propagates to Oliveiras equation (9) as follows.
	\begin{align}
		\dot{\gamma} &= f\left(\frac{2\lambda}{\eta_\text{p}}\left[\partial_xp\frac{r}{2^j} - \eta_\text{s}\dot{\gamma}\right]^2\right)\frac{1}{\eta_\text{p}}\left(\partial_xp\frac{r}{2^j}-\eta_\text{s}\dot{\gamma}\right)
		\intertext{This cannot be integrated as easily as without $\eta_\text{s}$, and is only solvable semi-analytically.}
		1 &= f\left(\frac{2\lambda}{\eta_\text{p}}\left[\partial_xp\frac{r}{2^j} - \eta_\text{s}\dot{\gamma}\right]^2\right)\frac{1}{\eta_\text{p}}\left(\partial_xp\frac{r}{2^j\dot{\gamma}}-\eta_\text{s}\right)
	\end{align}
	For the exponential version of PTT $f$ is defined as follows.
	\begin{equation}
		f\left(\Tr\ma{\tau}\right) = \exp\left(\frac{\epsilon\lambda}{\eta_\text{p}}\Tr\ma{\tau}\right)
	\end{equation}
	This can be inserted in the above equation.
	\begin{align}
		1 &= \exp\left(\frac{2\epsilon\lambda^2}{\eta_\text{p}^2}\left[\partial_xp\frac{r}{2^j} - \eta_\text{s}\dot{\gamma}\right]^2\right)\frac{1}{\eta_\text{p}}\left(\partial_xp\frac{r}{2^j\dot{\gamma}}-\eta_\text{s}\right) \\
		0 &= \frac{2\epsilon\lambda^2}{\eta_\text{p}^2}\left[\partial_xp\frac{r}{2^j} - \eta_\text{s}\dot{\gamma}\right]^2 - \ln\eta_\text{p} + \ln\left(\partial_xp\frac{r}{2^j\dot{\gamma}}-\eta_\text{s}\right) \\
		0 &= 2\epsilon\lambda^2\dot{\gamma}^2\left[\partial_xp\frac{r}{2^j\dot{\gamma}\eta_\text{p}} - \frac{\eta_\text{s}}{\eta_\text{p}}\right]^2 + \ln\left(\partial_xp\frac{r}{2^j\dot{\gamma}\eta_\text{p}}-\frac{\eta_\text{s}}{\eta_\text{p}}\right) \\
		0 &= 2\epsilon\lambda^2\dot{\gamma}^2\left[\partial_xp\frac{r}{2^j\dot{\gamma}\eta_\text{p}} - \frac{\eta_\text{s}}{\eta_\text{p}}\right]^2 + \ln\left(\partial_xp\frac{r}{2^j\dot{\gamma}\eta_\text{p}}-\frac{\eta_\text{s}}{\eta_\text{p}}\right)
		\intertext{This has the following form}
		0 &= ab^2 + \ln b
	\end{align}
	This can be solved using the Lambert W function as follows.
	\begin{align}
		b &= -\sqrt{\frac{W_0\left(2a\right)}{2a}}
		\intertext{Resolving for $r$ yields the following.}
		r &=  \frac{2^j\eta_\text{s}}{\partial_xp}\dot{\gamma} - \frac{\eta_\text{p}2^{j-1}}{\lambda\sqrt{\epsilon}\partial_xp}\sqrt{W_0\left(4\epsilon\lambda^2\dot{\gamma}^2\right)}
	\end{align}
	We notice, that this is the solution for a poiseuille flow with viscosity $\eta_\text{s}$ expanded by a term that applies a nonuniform scaling of the profile  along the $r$-axis towards higher radii. We simplify again using $c < 0$ and $d > 0$.
	\begin{equation}
		r = c\dot{\gamma} - R_\eta\frac{c}{d}\sqrt{W_0\left(d^2\dot{\gamma}^2\right)}
	\end{equation}

>>>>

To illustrate our novel LBM algorithm, we consider two prototypical cell carrier fluids.
The first is an alginate solution which is often used in biofabrication but can also serve as a common food additive.
Our solution contains \SI{4}{\percent} alginate dissolved in PBS and was measured in a commercial plate-plate rheometer (see appendix~\ref{app:experiment} for experimental details).
The shear-dependent viscosity $\eta$ and the first normal stress difference $N_1$ are shown as function of shear rate in figure~\ref{fig:fit_alginate}.
We note the very high zero-shear viscosity of more than \SI{10000}{\milli\pascal\second}.
Bioinks in biofabrication are in general characterized by such high viscosities to guarantee stability of the printed construct.
At the same time, during the actual printing process when the material is flowing, the viscosity should be as low as possible to avoid excessive mechanical stress on the embedded cells.
This motivates the use of highly shear-thinning viscoelastic fluids in biofabrication.
	\begin{figure}[H]
		\centering
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/fit_alginate_eta.pdf}
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/fit_alginate_N.pdf}
		\caption{Rheology data for alginate solution from 21 measurements of the viscosity (a) and normal stress (b).
				Orange lines show a fit of the PTT rheological model. \SG{ larger font in figures }\RK{done}}
		\label{fig:fit_alginate}
	\end{figure}

The second cell carrier fluid modeled in this work is a methyl cellulose solution which is commonly used in real-time-deformation-cytometry (RT-DC) experiments.
RT-DC experiments allow very high-throughput measurements of cell mechanical properties and have been used successfully for disease detection by screening blood samples.
We extract our data from \cite{buyukurganci2023} where the shear rheology of methyl cellulose was measured using a plate-plate and cone-plate rheometer.
Three concentrations of \SI{0.49}{\percent}, \SI{0.59}{\percent} and \SI{0.83}{\percent} methyl cellulose in PBS were investigated, and their rheology is shown in figure~\ref{fig:fit_mc}.
Even at the highest concentration the zero-shear viscosity is two orders of magnitude lower than for the considered alginate solution.
	\begin{figure}[H]
		\centering
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/fit_mc_eta.pdf}
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/fit_mc_N.pdf}
		\caption{The viscosity (a) and normal stress (b) of a methyl cellulose solution from \cite{buyukurganci2023}.
				Lines show a fit of the PTT rheological model. \SG{ include legend into figure. Maybe give color coding only in caption? }\RK{Done}\SG{ larger font in figures }\RK{done}}
		\label{fig:fit_mc}
	\end{figure}

@@

The PTT model is specified by a differential equation for the polymer stress $\ma{\tau}$:
	\begin{equation}
		\overset{\nabla}{\ma{\tau}} = -\xi\left(\ma{\tau}\cdot\ma{D} + \ma{D}\cdot\ma{\tau}\right) - e^{\frac{\epsilon\lambda}{\eta_\text{p}}\Tr\ma{\tau}}\frac{\ma{\tau}}{\lambda} + 2\frac{\eta_\text{p}}{\lambda}\ma{D}
		\label{eq:polymerStress}
	\end{equation}
where $\ma{D}$ is the strain rate tensor.
\SG{ check that PTT is given in this form in \cite{phan-thien1977,phan-thien1978} or cite other source }\RK{The given form is most similar to \cite{ferras2019}, which you removed?}
The polymer viscosity is denoted $\eta_\text{p}$ and $\lambda$ is the corresponding relaxation time.
$\epsilon$ is the so-called extensibility parameter which controls at what shear rates thinning takes place and $\xi$ describes the slip in the polymer network which we will set to zero in the following.
The upper convected derivative of an arbitrary tensor $\ma{A}$ is defined as follows:
	\begin{equation}
		\overset{\nabla}{\ma{A}} = \frac{\Dif\ma{A}}{\Dif t} - \left(\left(\nabla\vec{u}\right)^\mathrm{T}\cdot\ma{A} + \ma{A}\cdot\left(\nabla\vec{u}\right)\right)
	\end{equation}
where $\vec{u}$ is the fluid velocity and $\left(\nabla\vec{u}\right)_{ij} = \nabla_i\vec{u}_j$.
The material derivative is
	\begin{equation}
		\frac{\Dif\ma{A}}{\Dif t} = \frac{\partial\ma{A}}{\partial t} + \vec{u}\cdot\nabla\ma{A}.
	\end{equation}
Finally, the total fluid stress $\ma{\sigma}$ is the sum of the polymer stress $\ma{\tau}$ and a solvent stress due to the Newtonian solvent with viscosity $\eta_\text{s}$:
	\begin{align}
		\ma{\sigma} &= \ma{\tau} + 2\eta_\text{s}\ma{D} 		\label{eq:totalStress}
	\end{align}

@@

We introduce the ratio between the zero-shear and the infinite-shear (i.e.\ solvent) viscosity as\todo{nope $\eta_p$ is not zero shear viscosity!}
\begin{align}
	R_\eta = \frac{\eta_\text{p}}{\eta_\text{s}}		\label{eq:R_eta}
\end{align}
which will turn out to be an important parameter to determine numerical stability of LB simulations.
\SG{ wo don't actually show that $R_ \eta  $ determines stability? }\RK{we do not. We could do a plot with $\alpha_s$ decreasing until the sim breaks if you want}
This ratio can similarly be defined for other models such as FENE-P or the inelastic, but shear-thinning Carreau-Yasuda model \cite{carreau1972, yasuda1979}.
For our alginate solution we obtain $R_\eta = \num{48.2\pm0.4e3}$, indicating that the solution viscosity at very low shear rates is more than four orders of magnitude larger than the solvent viscosity which represents a severe challenge to standard numerical algorithms.
That this characteristic is not unique to alginate, but is in fact a generic feature of cell carrier fluids used in bioprinting is shown in table~\ref{tab:bioinks} where rheological data from various literature sources have been compiled into a single format to determine $R_\eta$.
The \SI{0.49}{\percent} methyl cellulose solution used in RT-DC measurements, in contrast, features a much lower value of $R_\eta = \num{18.7\pm0.4}$.

@@

The spatially resolved errors are shown in figure~\ref{fig:2DvelErr} in the appendix.
Averaging over the channel diameter, we obtain an L2 error of \num{9e-4} and \num{1e-3}\SG{ XX }\RK{done}\RK{we do not "average" over the diameter, we calculate the L2 error along the diameter} for alginate and methyl cellulose, respectively.
These conclusions remain robust even when varying the pressure gradient, and thus the flow velocity, over many orders of magnitude as shown in figure~\ref{fig:pressureErr}~(a) of the appendix.\RK{This is the worst case error for MC and it is tiny "remain robust" is too weak here}

	\begin{figure}[H]
		\centering
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/2DAlginateU.pdf}
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/2DParameterStudyUWorst.pdf}
		\caption{(a) Velocity profile of alginate in planar channel flow. \SG{use $y$ instead of $r$ in 2D}\RK{done}.
		The time evolution of the center-line velocity is shown in the inset \SG{ to do }.
		(b) Velocity profile for methyl cellulose.
		\SG{ make these plots for different $ \alpha _ \mathrm{s}$, either include here or put into appendix }}\RK{Why? Is identical down to fp32 precision (see appendix C)}
		\label{fig:2Dvel}
	\end{figure}\todo{maybe try if it is this identical for different $\alpha_s$ first}

We proceed to evaluate the viscosity shuffling method for a cylindrical Poiseuille flow.
The semi-analytical solution for this situation can be derived similarly as for the planar case and is given in appendix~\ref{app:semianalyticalPoiseuille}.
Numerically, the cylindrical situation is slightly more complex due to discretization errors ("staircase effect") when mapping the rounded channel walls onto the rectilinear LBM grid \cite{Lehmann.20229da}.
We use the same material parameters as above, but double the pressure gradient to \SG{ XX }. \SG{ why? } \RK{where analytical solutions exist in Poiseuille geometry the 3D solution differs from the 2D solution by a factor 2. See the velocities in our sims being the same with this scaling}
The channel has a diameter of \SI{20}{\micro\meter}\SG{ XX }\RK{done} and is discretized with 41 lattice points.
Again, we initialize the polymers in the unstretched state for methyl cellulose and in the stretched state for alginate.
The resulting velocity profile can be seen in figure~\ref{fig:3Dvel} and shows almost equally good agreement as for the planar case with an L2 error of \num{0.014} and \num{0.06}\SG{ XX }\RK{done}, respectively.
The spatially resolved error in figure~\ref{fig:3DvelErr} of the appendix shows that this larger average is indeed mainly attributable to the staircase effect as the error near the walls increases significantly.
The robustness against pressure gradient changes is shown in figure~\ref{fig:pressureErr}~(b) of the appendix.

	\begin{figure}[H]
		\centering
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/3DAlginateU.pdf}
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/3DParameterStudyUWorst.pdf}
		\caption{(a) Velocity profile of alginate in cylindrical channel flow.
		The time evolution of the center-line velocity when starting from a pre-stretched configuration (see main text) is shown in the inset.
		(b) Velocity profile for methyl cellulose.
				\SG{ make these plots for different $ \alpha _ \mathrm{s}$, either include or put into Si }}
		\label{fig:3Dvel}
	\end{figure}\todo{see above}

@@

	\begin{figure}[H]
		\centering
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/malaspinas2010.pdf}
		\caption{L2 error as defined in equation \ref{eq:l2error} from Malaspinas et al.\cite{malaspinas2010} (dots) compared to our algorithm (x) as a function of grid resolution.
		Blue and green crosses are not visible as they overlap exactly with the orange and red crosses, respectively, showing that the accuracy of our method is independent of $\mathrm{Wi}$ in the studied parameter range.}
		\label{fig:malaspinas2010}
	\end{figure}

As can be seen in figure~\ref{fig:malaspinas2010}, the accuracy of our method does not depend on the Weissenberg Number and is generally somewhat better than the method of \cite{malaspinas2010}.
We note a certain dependence of the accuracy on $R_\nu$ which we consider nevertheless acceptable given the overall smallness of the error.
Finally, we find that the error of our method decreases with lattice resolution to the power of \num{-2} \SG{ give number, show dashed line in plot }\RK{done}.\RK{They see diminishing returns we don't}

@@

The first application stems from bioprinting where the bioink is pushed through a conical nozzle \cite{possl2021, Poologasundarampillai.2021, oconnell2022}.
The nozzle radius is \SI{1.8}{\milli\meter} at the inlet and \SI{200}{\micro\meter} at the outlet.
The nozzle has a length of \SI{18}{\milli\meter}.
Due to the difference in inlet/outlet radii, the system cannot be modeled with periodic boundary conditions, and we use inflow/outflow boundary conditions as described in section~\ref{sec:Methods} instead.
The required values for velocity and polymer conformation are obtained from simulations of pipes of the respective radii and a fixed volume flow of $\approx\SI{3}{\micro\liter\per\second}$.
The resulting flow field can be seen in figure \ref{fig:Nozzle_v}~(a).
To quantify the total stress as a single number, we use the von Mises stress which is defined as \cite{Mueller.2023}:
\begin{align}
	\sigma_\text{vM} = \sqrt{\frac{1}{2}\left[\left(\sigma_1 - \sigma_2\right)^2 + \left(\sigma_2 - \sigma_3\right)^2 + \left(\sigma_3 - \sigma_1\right)^2\right]}.
\end{align}
It can be seen that the highest stresses arise at the nozzle exit in direct vicinity to the wall.
This means that even for fast flows, cells flowing close to the center experience less stress than cells flowing near the walls in agreement with earlier results in Newtonian fluids \cite{Mueller.2023}.
	\begin{figure}[H]
		\centering
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/Nozzle_v.pdf}
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/Nozzle_sigma_vM.pdf}
		\caption{Velocity profile (a) and von Mises stress (b) for methyl cellulose in a conical nozzle as used for bioprinting. \SG{ draw nozzle contour with black lines }}\RK{Done}
		\label{fig:Nozzle_v}
	\end{figure}

@@

 	\begin{figure}[H]
		\centering
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/RT-DC_v.pdf}
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/RT-DC_tau.pdf}
		\caption{Velocity profile (a) and shear stress (b) for methyl cellulose in a typical RT-DC geometry. \SG{ draw channel contour with black lines }
		\SG{ change to von Mises instead of $ \tau _{ 12 }$? }}
		\label{fig:RT-DC}
	\end{figure}

@@

	\begin{figure}[H]
		\centering
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/cellInShear_sigma_vM.pdf}
		\caption{The von Mises stress around a cell in viscoelastic shear flow.}
		\label{fig:cellInShear_vM}
	\end{figure}

@@

\section{Semi-analytical solution for Poiseuille flows of PTT fluids with solvent viscosity}
\label{app:semianalyticalPoiseuille}
\SG{ make a self-contained derivation which is understandable even without reading Oliveira \\}
\SG{ use Gl.~(\ref{eq:polymerStress}) as starting point and end with Poiseuille solution, but also shear solution Gl.~(\ref{eq:shearN}, \ref{eq:shearEta}) \\}
Our derivation is analogous to Oliveira\cite{oliveira1999}.

For both Poiseuille flow and shear flow the following holds.
\begin{equation}
	\vec{u} = \dot{\gamma}\left(y\right)\hat{e}_x
\end{equation}
With this the constitutive equation\todo{link} fo $\xi=0$ in steady state reduces to the following.
\begin{align}
	\exp\left(\frac{\epsilon\lambda}{\eta_\text{p}}\Tr\ma{\tau}\right)\tau_{xx} &= 2\lambda\dot{\gamma}\tau_{xy} \\
	\exp\left(\frac{\epsilon\lambda}{\eta_\text{p}}\Tr\ma{\tau}\right)\tau_{xy} &= \eta_\text{p}\dot{\gamma}
\end{align}
With $\Tr\ma{\tau} = \tau_{xy}$. In the three-dimensional case $y$ refers to the radial coordinate.
